Mapping short reads, faithfully

### 2020-02-10 manuscript.tex
Updates intro and abstract

\begin{abstract}
Mapping is the process of finding the original location of a DNA read in a
reference sequence, typically a genome. Short read mappers are used in
most applications of high-throughput sequencing. As such, they must be
continuously improved to keep up with increasing needs. To gain speed,
many mappers rely on seeding heuristics that can negatively affect the
mapping accuracy. For lack of a method to compute such probabilities,
mappers have so far used \textit{ad hoc} approximations of variable
quality. Here we devise the elements of a strategy to map short reads
faithfully, \textit{i.e.}, to provide accurate probabilities of mapping
errors. The key is to estimate the number of paralogus of the target so as
to spend more time on the reads that can be mapped with high confidence.
This strategy highlights the existence of a class of reads that can be
mapped with unprecedented confidence. We implement this strategy in a
prototype mapper that is competitive with state-of-the-art mappers BWA-MEM
and Bowtie2, with the benefit of faithfulness. The software is open-source
and availble from \url{https://github.com/gui11aume/smmfdp}.
\end{abstract}

@@

High throughput DNA sequencing is now a well-established technology with
countless applications in industry and medicine~\cite{pmid30084865}. The
Illumina short-read technology currently dominates the market of DNA
sequencing, so the software exploiting this technology is an important
target for optimization~\cite{Wikillu}.

After sequencing, reads are usually mapped to a known reference, typically
a genome, using a software tool known as a mapper. The main challenge is
that sequencers sometimes make mistakes, so mappers must solve an
approximate string-matching problem. The traditional vision of mapping
algorithms is to optimize speed and accuracy while maintaining a low
memory footprint. From this perspective, a major breakthrough was the
conception of the FM-index~\cite{ferragina2000opportunistic,
ferragina2005indexing}, a data structure based on the Burrows-Wheeler
transform~\cite{burrows1994block} and on the suffix
array~\cite{manber1993suffix}. Modern short-read mappers such as
BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286} rely on
specialized implementations of the FM-index for DNA sequences. More
recently, new mapping algorithms for specific applications such as RNA-seq
have further improved the speed and the accuracy in some
areas~\cite{pmid27043002}.

@@

Faithfulness comes into play for heuristic algorithms, \textit{i.e.},
algorithms that may return an incorrect result. We say that a heuristic is
\emph{faithful} if it provides an accurate probability that the result is
incorrect. Importantly, faithfulness is orthogonal to accuracy in the
sense that a faithful algorithm is not necessarily accurate and
\textit{vice versa}.

@@

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{pmid19505943}, featuring a mandatory \emph{mapping
quality} field MAPQ, defined as ``$-10 \cdot \log_{10}Pr$\{mapping
position is wrong\}''. However, the proposed methods to estimate this
probability are accurate only for some genomes and some sequencing
technologies.
THIS NEEDS TO BE WRITTEN BETTER.
%However, there is presently no way to estimate this
%probability, and therefore no way to respect the standard.
Some mappers
like STAR~\cite{pmid23104886} use the MAPQ field as a qualitative scale.
%Some others like GEM~\cite{pmid23103880} do not provide a mapping quality.
As for the mappers aiming to comply with the SAM standard, the scores are
undocumented, variable between mappers and versions, often spreading
puzzlement and frustration in the community~\cite{ACGT, QCFail}.

As a consequence, mapping quality is often neglected by users and
developpers, even though it is critical in many applications. For
instance, when calling \textit{de novo} mutations, the standard procedure
is to map sequencing reads to the genome of interest, and score mismatches
as evidence for mutations. Mapping errors can also explain the mismatches,
so the confidence in the final call is commensurate with the mapping
quality. Inaccuracies of the MAPQ estimates cause recurrent errors in
cancer genomics~\cite{dunno} where faithfulness would make a major
difference.

@@

Improving the mapping process requires to understand why it is sometimes
inaccurate in the first place. Mapping heuristics are usually based on a
filtration strategy called \emph{seeding}, whereby short matches between
the read and the reference are used to extract a set of candidate
locations. Seeding is error-prone because the correct location of the read
may be filtered out. We recently developped a formal computational
framework to compute the error rate of different seeding strategies used
for mapping short reads~\cite{Filion619155}. Missing from this framework
was a method to estimate the unknown parameters of the model.

Here we propose a method to estimate those parameters and we implement it
in a mapper called MEM Mapper Prototype (MMP) that aims to be faithful. We
use MMP to study the feasibility of this strategy and we highlight the
additional benefits gained from faithfulness. In the process, we discover
a class of reads that can be mapped with extremely high confidence and we
show that faithfulness can be achieved without sacrificing speed or
accuracy.

>>>>

\begin{abstract}
Mapping is the process of finding the original location of a DNA read in a
reference sequence, typically a genome. Short read mappers are software tools
used in most applications that involve high-throughput sequencing. As such, they
must be continuously improved to keep up with increasing needs. Modern mappers
rely on fast seeding heuristics that convert them in inexact algorithms.
For lack of a method to compute the reliability of reported mapping locations,
mappers have so far used \textit{ad hoc} approximations of variable
quality. Here we devise the elements of a strategy to map short reads
faithfully, \textit{i.e.}, to provide accurate probabilities of mapping reliability.
The key is to estimate the number of paralogs of the target reference sequence
so as to identify these reads that can be mapped with high confidence.
This approach highlights the existence of a class of reads that can be
mapped with unprecedented confidence. We exploit this strategy in a
prototype mapper that is competitive with state-of-the-art mappers BWA-MEM
and Bowtie2, with the benefit of faithfulness. The software is open-source
and availble at \url{https://github.com/gui11aume/smmfdp}.
\end{abstract}

@@

High throughput DNA sequencing is now a well-established technology with
countless applications in industry and medicine~\cite{pmid30084865}. The
Illumina short-read technology currently dominates the market of DNA
sequencing over newer technologies that promise longer reads at the
expense of higher error rates. Software and tools exploiting this technology
are therefore an important target for optimization~\cite{Wikillu}.

In most analyses, sequencing reads are mapped to a known reference, typically
a genome, using a software tool known as a mapper. Mappers are complex
algorithms that must solve approximate string-matching problems, provided that
the reads and the reference do not match perfectly due to biological divergence
andn due to measurement errors of sequencing machinery. The traditional vision
of mapping algorithms is to optimize speed and accuracy while maintaining a low
memory footprint. From this perspective, a major breakthrough was the
conception of the FM-index~\cite{ferragina2000opportunistic,
ferragina2005indexing}, a data structure based on the Burrows-Wheeler
transform~\cite{burrows1994block} and the suffix
array~\cite{manber1993suffix}. Modern short-read mappers such as
BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286} rely on
specialized implementations of the FM-index for DNA sequences. More
recently, new specialized algorithms such as RNA-seq mappers have further improved
the speed and the accuracy for their specific applications~\cite{pmid27043002}.

@@

Faithfulness comes into play for heuristic algorithms, \textit{i.e.},
algorithms that are not guaranteed to return a correct result. In this context,
a heuristic is \emph{faithful} if it provides an accurate probability that the
reported result is correct. Importantly, faithfulness is orthogonal to accuracy
in the sense that a faithful algorithm is not necessarily accurate and
\textit{vice versa}.

@@

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{pmid19505943}, featuring a mandatory \emph{mapping
quality} field MAPQ, defined as ``$-10 \cdot \log_{10}Pr$\{mapping
position is wrong\}''. However, there is presently no exact method to
estimate these probabilities and most mappers rely on approximations
that are accurate only for some specific genomes and sequencing technologies.
The different implementations of mapping quality estimators are usually
undocumented, variable between mappers and versions, often spreading
puzzlement and frustration in the community~\cite{ACGT, QCFail}.
Some mappers, like STAR~\cite{pmid23104886}, use the MAPQ field as a
qualitative scale.

As a consequence, mapping quality is often neglected by users and
developpers, even though it is critical in many applications. For
instance, when calling \textit{de novo} mutations, the standard procedure
is to map sequencing reads to the genome of interest and identify mismatches
as evidence for mutations. In this context, mapping errors cause artifacts that
explain nonexistent mismatches,
so the confidence in the final call is commensurate with the mapping
quality. Inaccuracies of the MAPQ estimates cause recurrent errors in
cancer genomics~\cite{dunno} where faithfulness would make a major
difference.

@@

Improving the mapping process requires to understand why it is sometimes
inaccurate in the first place. Mapping heuristics are usually based on a
filtration strategy called \emph{seeding}, whereby short exact matches between
the read and the reference are used to extract a set of candidate
locations. Seeding is error-prone because errors in the read may cause correct
location to be filtered out. We recently developped a formal computational
framework to compute the error rate of different seeding strategies used
for mapping short reads~\cite{Filion619155}.

Here we propose a method to estimate the unknown parameters of our previous
model and we implement it in a mapper called MEM Mapper Prototype (MMP) that
aims to be faithful. We use MMP to study the feasibility of this
strategy and we highlight the additional benefits gained from faithfulness.
In the process, we discover a class of reads that can be mapped with extremely
high confidence and we show that faithfulness can be achieved without sacrificing
speed or accuracy.
